---
title: "Xenium DCIS analysis"
output: html_document
author: Dr.Parul Kulshreshtha
---

Introduction

Background

In this analysis, I explore 10X Xenium spatial transcriptomics data from a tissue slide of breast ductal carcinoma in situ (DCIS), using data retrieved from Janesick, A. et al., High resolution mapping of the tumor microenvironment using integrated single-cell, spatial and in situ analysis. Nat Commun. (2023).

DCIS represents a form of pre-invasive breast carcinoma. In untreated cases, 25–60% of DCIS cases progress to invasive ductal carcinoma (IDC). A key clinical challenge lies in distinguishing between non-progressive and progressive DCIS, which often results in over- or under-treatment.

While DCIS is generally considered to have a better prognosis than IDC and is not regarded as a life-threatening disease, recurrence remains a concern. After treatment, the overall recurrence rate is approximately 20%, split evenly between in situ and invasive recurrence. Understanding the cellular and spatial dynamics of these lesions is therefore critical for improving patient outcomes.

Aims

In normal breast tissue, myoepithelial cells (MECs) play a key role in maintaining the structure and integrity of epithelial cells. In breast cancer, MECs act as a physical barrier against tumour cell invasion, and the integrity of the MEC layer has been associated with the risk of progression in DCIS patients.

In this analysis, I aim to:

1.Perform initial quality control (QC) on the Xenium dataset
2.Identify clusters of cells with similar gene expression profiles via unsupervised clustering
3.Assign cell type identities — including myoepithelial cells — to each cluster
4.Analyse the spatial distribution of myoepithelial cells in relation to other cell types


Xenium Data Loading and Initial Quality Checks (QC)

Data Structure
When working with spatial transcriptomic datasets, input data are typically organized across multiple files. For a Xenium dataset, the key components are:

Gene expression matrix: A cell × gene transcript count matrix, which also includes negative probes that can be used to estimate the signal-to-noise ratio of transcript quantification
Cell metadata table: A dataframe containing cell-level information such as cell identifiers, number of real transcripts and negative probes detected per cell, cell centroid x/y coordinates (µm), and total cell and nucleus area (µm²)
Polygon files: A list of x/y point coordinates used to reconstruct single-cell boundaries, useful for visualization and estimating cell-to-cell distances (e.g., membrane-to-membrane distance)
Transcript file: A list of x/y coordinates and cell assignments for each transcript detected within the Xenium sample

Loading the Data
R packages like Seurat provide helper functions to load all these files and organize their contents into a single object, using the LoadXenium() function. For this analysis, the Xenium sample has been pre-loaded and stored as a SeuratObject, which I access by loading the R object from ../data/Xenium_Breast_DCIS_SeuratObject.rds.
A SeuratObject is a specialized data structure from the Seurat R package that enables the study of gene expression in the context of tissue architecture.


```{r}
# Installing dependencies
all_pkgs = c("devtools", "remotes", "ggplot2", "ggsci", "ggpubr"
             , "viridis" , "DT", "plyr","dplyr"
             , "patchwork", "Matrix"
             , "Seurat", "msigdbr", "tidyverse"
             ,"Rfast", "magrittr", "stringr", "tibble", "dbscan"
             , "sf", "FNN"
)

lib = installed.packages()
installed = all_pkgs %in% rownames(lib)
not_installed = which(!installed)

if(length(not_installed)>0){
  for(i in not_installed){
    message("[*] Installing ", all_pkgs[i])
    if (all_pkgs[i] == "devtools"){
      install.packages("devtools")
    }
    if (all_pkgs[i] %in% all_pkgs){
      suppressMessages(suppressWarnings(install.packages(all_pkgs[i], dependencies = T)))
    }
  }
}

if(!require(devtools)) install.packages("devtools")

# Check installation completed successfully
lib = installed.packages()
installed = all_pkgs %in% rownames(lib)
not_installed = which(!installed)
message(length(not_installed), ' packages are not yet installed')
message(length(installed), ' out of ', length(all_pkgs), ' required packages are installed')
```

```{r}
# Plotting and visualization 
library(ggplot2)
library(ggpubr)
library(ggsci)

# Data manipulation 
library(plyr)
library(dplyr)
library(patchwork)
library(tidyverse)
library(magrittr)
library(Matrix)
#Spatial objects manipulation
library(sf)
library(FNN)

# Spatial transcriptomics analysis 
library(Seurat)
if("presto"%in% installed.packages()){library(presto)} # this package speeds up the computational time

# rmd rendering 
library(DT)
```

```{r}
# Load Seurat object containing pre-loaded Xenium data
xenium = readRDS('/Users/dr.parul/Downloads/Practical 4 - Xenium data analysis-20260323/data/Xenium_Breast_DCIS_SeuratObject.rds')
```

Initial Quality Checks (QC)
After loading the dataset, I begin by inspecting the Xenium single cells to get an overview of data quality. This involves examining three key metrics across cells:

Number of detected transcripts (nCount_Xenium): the total transcript count per cell
Number of unique genes (nFeature_Xenium): the number of distinct genes detected per cell
Number of negative probes (nCount_ControlProbe): sequences not expected to be found in human tissue, used to quantify the specificity of transcript detection

As with non-spatial single-cell RNA-seq data, the number of transcripts and genes helps identify low-quality cells — whether due to limited sequencing coverage or poor transcript detection efficiency. The lower the transcript and gene counts, the lower the quality.
A high negative probe count may indicate the presence of technical noise in the single-cell gene expression profiles. While negative probes are useful for this initial quality check, they are not used for cell phenotyping downstream.
I will use transcript and gene metrics to filter out low-quality cells before proceeding with downstream analyses, being careful not to apply overly stringent thresholds at this stage.
These three metrics are already stored in the SeuratObject as metadata columns:

nCount_Xenium
nFeature_Xenium
nCount_ControlProbe

```{r}
# Summary statistics
message('N genes = ', nrow(xenium))
message('N cells = ', ncol(xenium))

message('Median N transcripts per cell = ', median(xenium@meta.data$nCount_Xenium))
message('Median N genes per cell = ', median(xenium@meta.data$nFeature_Xenium))
message('Median N negative probes per cell = ', median(xenium@meta.data$nCount_ControlProbe))

# Plot number of transcripts per cell as a density plot
ggplot(xenium@meta.data, aes(x = nCount_Xenium, after_stat(count))) +
  theme_classic() +
  geom_density(fill = 'steelblue', color = NA, adjust = 0.5) +
  scale_x_continuous(breaks = seq(0, 1000, 100)) +
  ylab('N cells')
```

What the plot shows:

The distribution is right-skewed, which is completely normal for spatial transcriptomics data
The main peak is around 100 transcripts per cell, consistent with the median of 164 we saw earlier
There is a small bump near 0 on the left — these are the low-quality cells with very few transcripts that we'll want to filter out
The long tail to the right represents a small number of highly transcribed cells

For QC filtering, looking at this plot you can see there's a natural separation near the left bump. A common approach would be to set a minimum threshold around 10-20 transcripts per cell to remove those near-zero cells, without being too aggressive.
This would typically be done with something like:

```{r}
#Plot number of unique genes per cell as a density plot
ggplot(xenium@meta.data,aes(x=nFeature_Xenium,after_stat(count)))+
  theme_classic()+geom_density(fill='magenta',color=NA,adjust=0.5)+
  scale_x_continuous(breaks=seq(0,350,50))+ylab('N cells')
```


What the plot shows:

Main peak around 65-75 unique genes per cell — perfectly consistent with the median of 64 we saw earlier
The same small bump near 0 on the left — confirming these are the same low-quality cells with both low transcripts and low unique genes
Right-skewed tail tapering off around 150-200 genes, representing highly expressed cells
Most cells fall between 50-100 unique genes, which is expected given the 313-gene targeted panel

Comparing the two plots together:

Both plots show the same low-quality cell population near 0 — this gives us more confidence that these are genuinely poor quality cells
The shapes are very similar, which is expected since transcript count and gene count are usually correlated
The gene plot has a tighter, more defined peak than the transcript plot, which is normal.



```{r}
#Plot number of negative probes per cell as a density plot
ggplot(xenium@meta.data,aes(x=nCount_ControlProbe,after_stat(count)))+
  theme_classic()+geom_density(fill='firebrick3',color=NA,adjust=5)+
  scale_x_continuous(breaks=seq(0,10,2))+ylab('N cells')
```

What the plot shows:

The dominant peak at 0 — the vast majority of cells (~200,000+) have zero negative probes detected
A very small secondary bump around 1 — a tiny fraction of cells with 1 negative probe
An even smaller bump around 2 — negligible number of cells
Essentially nothing beyond 4-5 negative probes

What this means:

The dataset has very low technical noise — confirming what the median of 0 already told us
The near-absence of negative probes suggests the transcript detection is highly specific
This is a high quality Xenium run — you won't need aggressive filtering based on negative probes

All three QC plots together tell a consistent story:

1.Reasonable transcript counts (median 164)
2.Good unique gene detection (median 64)
3.Virtually no technical noise (median 0 negative probes)

```{r}
# Apply a first permissive filtering step to Xenium cell based on total transcripts/genes and signal-to-noise ratio (i.e., (total NegativeProbe counts)/(total transcript counts))
xenium_cells = xenium@meta.data
message('[*] Removing cells with less than 20 transcripts')
xenium_cells = xenium_cells %>% filter(nCount_Xenium >= 20)

message('[*] Removing cells with less than 15 total expressed genes')
xenium_cells = xenium_cells %>% filter(nFeature_Xenium >= 15)

message('[*] Removing cells with a ratio of negative probes/real transcripts >= 5%')
xenium_cells = xenium_cells %>% filter(nCount_ControlProbe/nCount_Xenium < 0.05)

message('Cells passed all filtering steps: ', nrow(xenium_cells),' out of ', nrow(xenium@meta.data))

# Removing cells from Xenium object
xenium = UpdateSeuratObject(xenium)
xenium = subset(xenium, cells = rownames(xenium_cells))
rm(xenium_cells)

# Set Xenium cell ids as the main cell identifier in SeuratObject
Idents(xenium) = 'cell_id'
```

1. How many cells are included in the dataset?

The dataset contains 167,780 cells before filtering, as shown by ncol(xenium)

2. How many genes are included in the Xenium panel?

The Xenium panel includes 313 genes, as shown by nrow(xenium)

3. What is the median number of transcripts, genes, and negative probes across cells?


Median transcripts per cell: 164 (nCount_Xenium)
Median unique genes per cell: 64 (nFeature_Xenium)
Median negative probes per cell: 0 (nCount_ControlProbe)


4. How many cells do you retain after applying the quality filtering steps?

161,789 cells are retained after applying the three filtering steps:

Removing cells with < 20 transcripts
Removing cells with < 15 unique genes
Removing cells with a negative probe/transcript ratio ≥ 5%

This means 5,991 cells (~3.5%) were removed, indicating the dataset was already of high quality.


Data Preprocessing: Normalisation, Feature Selection and Sketching

Before proceeding to clustering and downstream analysis, I perform three key preprocessing steps:

1. Data Normalisation

Raw single-cell counts are heavily influenced by technical factors such as sequencing depth (library size) and capture efficiency. Without normalisation, cells with higher read counts would artificially appear to have higher expression across most genes. Normalising the data makes gene expression values comparable between cells, reducing technical variability while preserving the underlying biological signal.

2. Variable Feature Selection

Not all genes are equally informative. Many genes show little variation across cells — such as housekeeping genes — while others capture true biological differences, such as cell-type markers. I therefore identify the most variable features (genes), focusing on those with high biological variability relative to technical noise. This ensures that downstream clustering is driven by the most informative genes.

3. Data Sketching

Modern single-cell datasets can contain hundreds of thousands to millions of cells, making full analysis memory-intensive and slow. To address this, I apply data sketching, which selects a representative subset of cells and performs dimensionality reduction and clustering on this subset. The resulting clusters are then projected back onto the full dataset.

More information about data sketching in single-cell analysis can be found on the Seurat website.

```{r}
# --- Normalisation ---
# Log-normalise count data
xenium <- NormalizeData(xenium, assay = "Xenium", 
                        normalization.method = "LogNormalize",
                        results = 'hide')

# --- Variable Feature Selection ---
# Identify top variable genes using variance stabilizing transformation (vst)
# Fits log(variance) ~ log(mean) using loess regression, then standardizes 
# feature values using observed mean and expected variance from the fitted line
xenium <- FindVariableFeatures(xenium, selection.method = "vst", 
                               nfeatures = nrow(xenium))

# --- Data Sketching ---
# Select a representative subset of 75,000 cells using Leverage Score sampling
xenium <- SketchData(object = xenium,
                     ncells = 75000,
                     method = "LeverageScore",
                     sketched.assay = "sketch")

# Set sketched data as default assay for clustering
DefaultAssay(xenium) <- "sketch"

# Repeat variable feature selection on sketched data
xenium <- FindVariableFeatures(xenium)

# Scale and centre features
xenium <- ScaleData(xenium)

# --- Dimensionality Reduction ---
# Perform PCA computing 50 principal components
xenium <- RunPCA(xenium, reduction.name = "pca", npcs = 50)

# --- Clustering ---
# Build nearest neighbour graph using all 50 PCs
xenium <- FindNeighbors(xenium, reduction = "pca", dims = 1:50)

# Identify clusters using Louvain algorithm (resolution = 0.8)
# Higher resolution = more clusters
xenium <- FindClusters(xenium, cluster.name = "seurat_clusters", 
                       resolution = 0.8)

# --- UMAP Visualisation ---
# Generate UMAP embedding using all 50 PCs
# return.model = TRUE allows projection of remaining cells later
xenium <- RunUMAP(xenium, dims = 1:50, return.model = TRUE)

# --- Project Clusters to Full Dataset ---
# Extend cluster annotations from sketched subset to all remaining cells
xenium <- ProjectData(object = xenium,
                      assay = "Xenium",
                      full.reduction = "pca.full",
                      sketched.assay = "sketch",
                      sketched.reduction = "pca",
                      umap.model = "umap",
                      dims = 1:50,
                      refdata = list(cluster_full = "seurat_clusters"))

# --- Finalise ---
# Set full Xenium dataset as default assay and cluster_full as active identity
DefaultAssay(xenium) <- "Xenium"
Idents(xenium) <- 'cluster_full'
message(length(unique(xenium$cluster_full)), ' cell clusters identified.')

# Visualise clusters over spatial image
ImageDimPlot(xenium, axes = TRUE)


```

Cluster Annotation: Identifying Cell Type Identities

Differential Expression Analysis

Having divided cells into 32 clusters in the previous section, I now aim to assign a biological identity to each cluster — that is, to determine which cell types are present within each cluster.

To do this, I perform differential expression (DE) analysis, comparing cells within each cluster against all remaining cells in the dataset. This produces a list of cluster-specific marker genes that best distinguish each cluster from the rest. Genes that are significantly upregulated in a given cluster relative to all others are considered that cluster's markers.

Cell Type Annotation

Once I have a set of specific genes for each cluster, I intersect these with well-known literature-based cell type markers. For each cluster, I search for a cell type whose known markers are over-represented — or enriched — in the list of cluster-specific genes. This allows me to assign a putative cell type identity to each cluster.
To speed up the annotation process, I use a pre-made list of cell type markers, loaded from:

```{r}
# Load pre-made cell type markers
cell_type_markers <- readRDS('/Users/dr.parul/Downloads/Practical 4 - Xenium data analysis-20260323/data/Xenium_Breast_DCIS_cellType_markers.rds')
```

```{r}
# Set full Xenium dataset as default assay
DefaultAssay(xenium) <- "Xenium"

# --- Differential Expression Analysis ---
# Find positive marker genes for each cluster using Wilcoxon rank-sum test
# only.pos = TRUE: returns only upregulated markers (FC > 0)
# min.pct = 0.01: only test genes detected in at least 1% of cells
# logfc.threshold = 0.1: minimum log-fold change threshold
markers <- FindAllMarkers(xenium, only.pos = TRUE,
                          min.pct = 0.01, logfc.threshold = 0.1, 
                          test.use = "wilcox")

# Rank markers within each cluster by adjusted p-value
markers <- ddply(markers, .(cluster), mutate, rank = rank(p_val_adj))

# --- Marker Filtering ---
# Filter markers: log2FC >= log2(1.5) and adjusted p-value <= 0.05
markers %>%
  group_by(cluster) %>%
  dplyr::filter(avg_log2FC >= log2(1.5) & p_val_adj <= 0.05) -> selected_markers

# Select top 10 and top 5 marker genes per cluster
selected_markers %>% slice_head(n = 10) %>% ungroup() -> top10
selected_markers %>% slice_head(n = 5) %>% ungroup() -> top5

# --- Visualisation ---
# Dot plot of top 5 marker genes per cluster
# dot.min = 0.2: minimum dot size threshold
DotPlot(xenium, features = unique(top5$gene), dot.min = 0.2, 
        cluster.idents = F) +
  theme(axis.text.x = element_text(angle = 90, size = 5, 
                                   hjust = 0.95, vjust = 0.2))
```

Key observations:

There is a clear diagonal pattern of high expression — this means each cluster has a distinct set of highly specific marker genes, which is a sign of good clustering quality
Some clusters like 17, 26, 29, 3 (bottom left) show very bright blue, large dots — these have very specific and highly expressed markers, likely representing well-defined cell types
Clusters 13, 10, 28 (top) show high expression across many genes on the right side — these may represent transcriptionally active cell types
Some clusters show broader, more diffuse expression patterns, suggesting they may be related cell types or transitional states

```{r}
# --- Cell Type Annotation ---
# Load pre-made cell type marker genes and merge with top 10 cluster markers
xenium_gene_groups <- readRDS('/Users/dr.parul/Downloads/Practical 4 - Xenium data analysis-20260323/data/Xenium_Breast_DCIS_cellType_markers.rds')
cluster_ctype <- merge(top10, xenium_gene_groups, by = 'gene') %>% 
  select(cluster, gene, cellType)

# Assign each cluster to a cell type based on maximum marker gene overlap
cluster_ctype <- cluster_ctype %>% 
  dplyr::count(cluster, cellType) %>% 
  dplyr::group_by(cluster) %>% 
  dplyr::mutate(prop = n/sum(n)) %>%
  dplyr::filter(prop == max(prop)) %>% 
  dplyr::ungroup()

# If a cluster maps to multiple cell types, concatenate them
cluster_ctype <- ddply(cluster_ctype, .(cluster), function(x){
  x$cellType_anno <- ifelse(nrow(x) == 1, x$cellType, 
                            paste0(x$cellType, collapse = '.'))
  unique(x[, c('cluster', 'cellType_anno')])
})

# --- Add Cell Type Annotation to Seurat Metadata ---
xenium$cluster_full <- as.character(xenium$cluster_full)
cluster_ctype$cluster <- as.character(cluster_ctype$cluster)

meta <- merge(xenium@meta.data, cluster_ctype, 
              by.x = 'cluster_full', by.y = 'cluster')
rownames(meta) <- meta$cell_id
xenium@meta.data <- meta[colnames(xenium), ]

Idents(xenium) <- 'cellType_anno'

# --- Define Colour Palette ---
my_palette <- c("dodgerblue2", "#E31A1C", "green4", "#6A3D9A", "skyblue2",
                "#FF7F00", "palegreen2", "gold1", "#FB9A99", "#CAB2D6", 
                "#FDBF6F", "gray70", "khaki2", "maroon", "orchid1", 
                "deeppink1", "blue1", "steelblue4", "darkturquoise", 
                "green1", "yellow4", "yellow3", "darkorange4", "brown")

my_palette <- my_palette[1:length(levels(Idents(xenium)))]
my_palette <- setNames(object = my_palette, levels(Idents(xenium)))

# --- Visualise Cell Types on UMAP ---
DimPlot(xenium, cols = my_palette)
```

```{r}
# --- Cell Type Abundance ---
# Rank cell types by total cell count across the Xenium sample
ctype_freqs <- xenium@meta.data %>% 
  dplyr::count(cellType_anno) %>% 
  arrange(desc(n))

# Extract top 3 most abundant cell types
top3_ctypes <- ctype_freqs$cellType_anno[1:3]
message('Top 3 most abundant cell types:\n', paste0(top3_ctypes, collapse = '\n'))

# Set cell type order by abundance for plotting
ctype_freqs$cellType_anno <- factor(ctype_freqs$cellType_anno, 
                                    levels = ctype_freqs$cellType_anno)

# --- Visualisation ---
# Bar plot of total cell count per cell type
ggplot(ctype_freqs, aes(x = cellType_anno, y = n, label = n)) + 
  theme_classic() +
  geom_bar(stat = 'identity', fill = 'steelblue') +
  theme(axis.text.x = element_text(angle = 90, size = 10, 
                                   hjust = 0.95, vjust = 0.2)) +
  geom_text(vjust = -0.25)
```
Top 3 most abundant cell types:

Glandular cells — makes complete sense as the dominant cell type since DCIS originates in the ductal/glandular epithelium of the breast. These are the normal epithelial cells lining the ducts.
T cells — a large immune infiltrate is expected in DCIS, as the immune system actively responds to the developing tumour. A high T cell presence is consistent with an active anti-tumour immune response.
Fibroblasts — very common in breast tumour microenvironments, where they form the stromal scaffold around the tumour. They are known to play a role in tumour progression.



Myoepithelial cells are clearly present but not in the top 3, which is expected — they form a single protective layer around the ducts rather than filling the tissue
The dominance of Glandular cells over Cancer cells suggests the tissue contains a mix of normal and tumour ductal cells, which is typical for DCIS


```{r}
# --- Spatial Visualisation ---
# Visualise cells in tissue space using centroid coordinates
ImageDimPlot(xenium, axes = T, cols = my_palette)
```

```{r}
# Visualise cells using whole segmented cell boundaries
DefaultBoundary(xenium[["fov"]]) <- 'segmentation'
ImageDimPlot(xenium, axes = T, cols = my_palette, border.color = NA)
```

Spatial Proximity Analysis

By examining the spatial plots generated in the previous section, I can observe that myoepithelial cells tend to organise as a monolayer around duct-like structures formed by glandular cells — consistent with their known biological role as a protective barrier in normal breast tissue.

To further explore this spatial relationship, I calculate the distances between myoepithelial cells and cells belonging to all other cell types. Specifically, for each myoepithelial cell, I identify the closest cell of each other cell type and record that minimum distance. This gives a distribution of minimum distances between myoepithelial cells and each other cell type.

I then compare these distance distributions across cell types using a Wilcoxon rank-sum test, which allows me to statistically assess whether myoepithelial cells are significantly closer to glandular cells than to other cell populations. If the myoepithelial-glandular distance distribution is significantly smaller than the distances to other cell types, this would provide quantitative evidence for the spatial association between these two cell types — and by extension, support the hypothesis that myoepithelial cells form a protective layer around glandular/ductal structures in DCIS tissue.


```{r}
# --- Spatial Proximity Analysis ---
# Split cell metadata into myoepithelial cells and all other cell types
cell_coords_myo <- xenium@meta.data %>% 
  select(cell_id, cellType_anno, cluster_full, x_centroid, y_centroid) %>% 
  filter(cellType_anno == 'Myoepithelial_cells')

cell_coords_other <- xenium@meta.data %>% 
  select(cell_id, cellType_anno, cluster_full, x_centroid, y_centroid) %>% 
  filter(cellType_anno != 'Myoepithelial_cells')

# --- K-Nearest Neighbour Distance Calculation ---
# For each myoepithelial cell, calculate the minimum Euclidean distance 
# to the nearest cell of each other cell type using centroid coordinates
cell_coords_other <- split(cell_coords_other, cell_coords_other$cellType_anno)

myo_knn <- lapply(cell_coords_other, function(x){
  knn <- get.knnx(data = x[, c('x_centroid', 'y_centroid')],
                  query = cell_coords_myo[, c('x_centroid', 'y_centroid')],
                  k = 1)$nn.dist
  data.frame(myo_id = cell_coords_myo$cell_id, 
             knn_cellType = unique(x$cellType_anno), 
             knn_dist = knn[, 1])
})

# Bind results across all cell types
myo_knn <- do.call('rbind', myo_knn)

# --- Rank Cell Types by Median Minimum Distance ---
ranked_knn <- myo_knn %>% 
  group_by(knn_cellType) %>% 
  reframe(dist = median(knn_dist)) %>%
  arrange(dist)
ranked_knn

# Set cell type order by distance for plotting
myo_knn$knn_cellType <- factor(myo_knn$knn_cellType, 
                                levels = ranked_knn$knn_cellType)

# --- Visualisation ---
# Boxplot of minimum distances from myoepithelial cells to each cell type
# Wilcoxon test comparing distance to Glandular_cells vs Fibroblasts
ggplot(myo_knn, aes(x = knn_cellType, y = knn_dist, fill = knn_cellType)) +
  theme_classic() +
  geom_boxplot(outlier.shape = NA) +
  scale_fill_manual(values = my_palette) +
  labs(x = 'Cell type', 
       y = 'Distance from myoepithelial cells (um)') +
  theme(axis.text.x = element_text(angle = 90, size = 10, 
                                   hjust = 0.95, vjust = 0.2)) +
  stat_summary(fun = "median", geom = "label", 
               aes(label = round(after_stat(y))), fill = 'white') +
  ggpubr::stat_compare_means(comparisons = list(c("Glandular_cells", "Fibroblasts")),
                             method = "wilcox.test", label.y = 300)
```


Conclusion

In this analysis, I performed a comprehensive spatial transcriptomics analysis of a breast ductal carcinoma in situ (DCIS) tissue sample using the 10X Xenium platform.
Following quality control, I retained 161,789 out of 167,780 cells, confirming the high quality of the dataset. Unsupervised clustering identified 32 distinct cell clusters, which were subsequently annotated into 15 cell types based on their marker gene expression profiles. The three most abundant cell types were Glandular cells, T cells, and Fibroblasts, reflecting the complex cellular composition of the DCIS tumour microenvironment.

Spatial proximity analysis revealed that Myoepithelial cells are significantly closer to Glandular cells than to any other cell type, with a median minimum distance of 13.39 µm — nearly half the distance to the second nearest cell type, Fibroblasts (24.07 µm). The Wilcoxon rank-sum test confirmed this difference to be statistically significant. These findings are consistent with the known biological role of myoepithelial cells as a protective monolayer surrounding ductal/glandular structures in breast tissue.

Taken together, these results demonstrate the power of spatial transcriptomics in capturing not only the transcriptional identity of individual cells, but also their spatial organisation within the tissue. The quantitative spatial association between myoepithelial and glandular cells provides a basis for further investigation into how disruption of the myoepithelial layer may contribute to the progression of DCIS towards invasive ductal carcinoma.


